---
title: "Plot Principal Component Analysis (PCA) of 1kGP"
author: "Ahmed Moustafa"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M %z')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries

```{r load-packages, message=FALSE}
library(tidyverse)
library(tidylog)
library(printr)
library(Rtsne)
library(umap)
```

## Data Loading

### Metadata (populations information)

```{r load-samples-metadata}
samples = read_tsv("data/populations.tsv")
head(samples)
```

### Eigenvalues

```{r load-eigenvalues}
# Load the eigenvalues
eigenval = read_tsv("working/1kGP_pca.eigenval", col_names = c("value"))
head(eigenval)
```

### Eigenvectors

```{r load-eigenvectors}
# Load the eigenvectors
eigenvec = read.table("working/1kGP_pca.eigenvec", sep = " ")
dim(eigenvec)
```

```{r}
# Display the first few rows of the PCA data to inspect the structure
head(eigenvec)
```

```{r extract-pcs}
pcs = eigenvec[, 3:ncol(eigenvec)]
colnames(pcs) = paste0("PC", 1:ncol(pcs))
rownames(pcs) = eigenvec$V1
head(pcs)
```

```{r create-pca-dataframe}
pca_df = data.frame(sample = row.names(pcs), pcs) %>% inner_join(samples)
head(pca_df)
```

## Variance Explained

```{r calculate-variance-explained-by-each-component}
variance_explained = tibble(pc = paste0("PC", 1:nrow(eigenval)), var = eigenval$value / sum(eigenval))
head(variance_explained)
```

### Plotting the Variance Explained

```{r}
ggplot(variance_explained %>% top_n(3)) + 
  geom_bar(aes(x = pc, y = var), stat = "identity") + 
  theme_light() +
  labs (x = "PC", y = "Variance Explained") +
  scale_y_continuous(labels = scales::percent)
```



## Visualizing PCA Clustering

### By population

```{r plot-pca-by-population}
ggplot(pca_df) + 
  geom_jitter(aes(x = PC1, y = PC2, color = population), alpha = 0.7) + 
  theme_light()
```

### By superpopulation

```{r plot-pca-by-superpopulation}
ggplot(pca_df) + 
  geom_jitter(aes(x = PC1, y = PC2, color = superpopulation), alpha = 0.7) + 
  theme_light()
```

## Alternative Clustering using t-SNE

### Perform t-SNE

```{r perform-tsne}
# Perform t-SNE
set.seed(123)  # Set a seed for reproducibility
tsne_result = Rtsne(as.matrix(pcs[,1:3]), dims = 2, perplexity = 30, theta = 0.5, verbose = TRUE)

# The results are stored in tsne_result$Y
head(tsne_result$Y)
```

### Visualizing t-SNE Clustering

```{r create-tsne-dataframe}
# Create a data frame for plotting
tsne_df = data.frame(TSNE1 = tsne_result$Y[, 1], 
                     TSNE2 = tsne_result$Y[, 2], 
                     sample = rownames(pcs)) %>% inner_join(samples)
head(tsne_df)
```
#### By population

```{r plot-tsne-by-population}
ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = population), alpha = 0.7) +
  geom_point() +
  theme_light() +
  labs(x = "t-SNE1", y = "t-SNE2")
```
#### By superpopulation

```{r plot-tsne-by-superpopulation}
ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = superpopulation), alpha = 0.7) +
  geom_point() +
  theme_light() +
  labs(x = "t-SNE1", y = "t-SNE2")
```
## Alternative Clustering using UMAP

### Setting UMAP parameters

```{r}
# Set parameters for UMAP
umap_config = umap::umap.defaults
umap_config$n_neighbors = 20
umap_config$min_dist = 0.9
umap_config$metric = "euclidean"
```

### Perform UMAP

```{r}
# Perform UMAP
set.seed(123)  # Set a seed for reproducibility
umap_result = umap(as.matrix(pcs[, 1:3]), config = umap_config)

# The results are stored in umap_result$layout
head(umap_result$layout)
```

### Visualizing UMAP Clustering

```{r}
# Create a data frame for plotting
umap_df = data.frame(UMAP1 = umap_result$layout[, 1], 
                     UMAP2 = umap_result$layout[, 2], 
                     sample = rownames(pcs)) %>% inner_join(samples)
head(umap_df)
```

#### By population

```{r plot-umap-by-population}
# Plot the UMAP results
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = population), alpha = 0.7) +
  geom_point() +
  theme_light() +
  labs(x = "UMAP1", y = "UMAP2")
```

#### By superpopulation

```{r plot-umap-by-superpopulation}
# Plot the UMAP results
ggplot(umap_df, aes(x = UMAP1, y = UMAP2, color = superpopulation), alpha = 0.7) +
  geom_point() +
  theme_light() +
  labs(x = "UMAP1", y = "UMAP2")
```


## Further Readings

- [Dimensionality Reduction: PCA, t-SNE, and UMAP](https://medium.com/@aastha.code/dimensionality-reduction-pca-t-sne-and-umap-41d499da2df2)
- [Seeing data as t-SNE and UMAP do](https://pubmed.ncbi.nlm.nih.gov/38789649/) [[pdf](https://www.nature.com/articles/s41592-024-02301-x.pdf)]


## Session Info

```{r display-session-info}
sessionInfo()
```
