---
title: "Plot Principal Component Analysis (PCA) of 1kGP"
author: "Ahmed Moustafa"
date: "`r format(Sys.time(), '%d %B, %Y %H:%M %z')`"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading Libraries

```{r load-packages, message=FALSE}
library(tidyverse)
library(tidylog)
library(Rtsne)
```

## Data Loading: Eigenvalues

```{r load-eigenvalues}
# Load the eigenvalues
eigenval = read_tsv("working/1kGP_pca.eigenval", col_names = c("value"))
head(eigenval)
```

## Calculating Variance Explained by Each Principal Component

```{r}
variance_explained = tibble(pc = paste0("PC", 1:nrow(eigenval)), var = eigenval$value / sum(eigenval))
head(variance_explained)
```

## Plotting the Variance Explained by Principal Components

```{r}
ggplot(variance_explained %>% top_n(3)) + 
  geom_bar(aes(x = pc, y = var), stat = "identity") + 
  theme_light() +
  labs (x = "PC", y = "Variance Explained") +
  scale_y_continuous(labels = scales::percent)
```

## Data Loading: Eigenvectors

```{r load-eigenvectors}
# Load the eigenvectors
eigenvec = read.table("working/1kGP_pca.eigenvec", sep = " ")
dim(eigenvec)
```

```{r}
# Display the first few rows of the PCA data to inspect the structure
head(eigenvec)
```

```{r}
pca = tibble(sample = eigenvec$V1, PC1 = eigenvec$V3, PC2 = eigenvec$V4, PC3 = eigenvec$V5)
head(pca)
```

## Plotting PCA: Visualization of Population Structure

```{r}
ggplot(pca) + 
  geom_jitter(aes(x = PC1, y = PC2), alpha = 0.7) + 
  theme_light()
```

## Enhancing the PCA Plot: Color by Population

```{r load-samples-metadata}
samples = read_tsv("data/populations.tsv")
head(samples)
```
```{r}
pca2 = pca %>% inner_join(samples)
head(pca2)
```
```{r}
ggplot(pca2) + 
  geom_jitter(aes(x = PC1, y = PC2, color = superpopulation), alpha = 0.7) + 
  theme_light()
```

```{r}
ggplot(pca2) + 
  geom_jitter(aes(x = PC1, y = PC2, color = population), alpha = 0.7) + 
  theme_light()
```


## Alternative Clustering using t-SNE

```{r}
n = 10
df = data.frame (eigenvec[, 3:(n+2)])
colnames(df) = paste0("PC", 1:n)
rownames(df) = eigenvec$V1
head(df)
```

### Perform t-SNE

```{r perform-tsne}
# Perform t-SNE
set.seed(123)  # Set a seed for reproducibility
tsne_result <- Rtsne(as.matrix(df), dims = 2, perplexity = 30, theta = 0.5, verbose = TRUE)

# The results are stored in tsne_result$Y
head(tsne_result$Y)

```

### Visualizing t-SNE Clustering

```{r create-tsne-dataframe}
# Create a data frame for plotting
tsne_df = data.frame(TSNE1 = tsne_result$Y[, 1], 
                     TSNE2 = tsne_result$Y[, 2], 
                     sample = rownames(df)) %>% inner_join(samples)
head(tsne_df)
```

```{r plot-tsne-by-population}
ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = population), alpha = 0.7) +
  geom_point() +
  theme_light()
  labs(x = "t-SNE1", y = "t-SNE2")
```

```{r plot-tsne-by-superpopulation}
ggplot(tsne_df, aes(x = TSNE1, y = TSNE2, color = superpopulation), alpha = 0.7) +
  geom_point() +
  theme_light()
  labs(x = "t-SNE1", y = "t-SNE2")
```

## Session Info

```{r display-session-info}
sessionInfo()
```
